---
title: "Report for `r params$country`"
author: "Amit Levinson"
date: '`r format(Sys.Date(), "%d %B, %Y")`'
output: pdf_document
params: 
  continent: "Africa"
  country: "Morocco"
  year: 2007
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 5, fig.height = 3)
```

This an exmaple Rmarkdown file to show how we can easily render reports for many parameters across a variable.

We'll use for the example the dataset `gapminder` which includes information about countries' population, life expectancy and GDP across several years. This is a sample from the original gapminder but will do for the current exmaple.

We'll start with loading our packages:

```{r echo = TRUE}
library(gapminder)
library(tidyverse)
library(glue)
theme_set(theme_minimal())
```

```{r echo = FALSE}
 #unique(gapminder$country)
# gapminder::gapminder %>% filter(country == "Morocco")
```


Here we can add more text with different features such as:

- *italic*

- **bold**

a list:

1.

2.


# H1

## H2

### H3

<hr/>

## Descriptives

We first want to look at some descriptive to understand where our country is located in comparison to others.


Therefore let's look at `r params$country` life expectancy across years in comparison to other countries:

```{r fig.cap= "Here we can add a nice caption explaining a little more about the graph", fig.height=2.5}
ggplot(gapminder, aes(x = year, y= lifeExp, group = country))+
  geom_line(color = "gray75", alpha = 0.6)+
  geom_line(data = subset(gapminder, country == params$country), color = "black", alpha = 1, size = 0.6)+
  geom_text(data = subset(gapminder, country == params$country & year == max(year)), aes(label = country), hjust = 1, vjust = 1.1)+
  coord_cartesian(clip = "off")+
  scale_x_continuous(limits = c(1950, 2007))
```


### Or we can look at the average with 1 sd each way:

```{r fig.height = 6, fig.width=5}
gapminder %>% 
  filter(continent == params$continent) %>% 
  group_by(country) %>%
  summarise_at(vars(gdpPercap, lifeExp, pop), list(mean = mean, sd = sd), na.rm = T) %>% 
  ungroup() %>%
  mutate(is_country = ifelse(country == params$country, "y", "n")) %>% 
  ggplot(aes(x = pop_mean, y = reorder(country, pop_mean), color = is_country))+
  geom_point()+
  geom_errorbar(aes(xmin = pop_mean - pop_sd, xmax = pop_mean + pop_sd), width = .2)+
  scale_x_log10(labels = scales::comma)+
  labs(x = "Population mean (log scale)", title = glue("Average population from 1952-2007 for {params$continent}"))+
  scale_color_manual(values = c(y = "black", n = "gray75"))+
  theme(
    legend.position = "none"
  )
```


Another thing we might be interested in is `r params$country`'s life expectancy and GDP per capita for the year `r params$year`:

```{r fig.align='center'}
gapminder_year <- gapminder %>% 
  filter(year == params$year)
  
ggplot(gapminder_year, aes(x = gdpPercap, y = lifeExp))+
  geom_point(alpha = 0.6, color = "gray75")+
  geom_point(data = subset(gapminder_year, country == params$country),
             alpha = 1, color = "black")+
  geom_text(data = subset(gapminder_year, country == params$country),
            label= params$country, vjust = -0.5)+
  scale_x_log10()+
  theme_minimal()+
  labs(title = glue("Relationship between life expectancy and GDP \nper capita for {params$year}"), x = "GDP per capita", y = "Life expectancy")+
  theme(plot.title =element_text(size = 10))
```


## Regression Analysis

And of course we can easily run a regression and beautifuly print it to a table. Again, we can use the `params$` feature we created in the beginning to chose the country we need and print the table accordingly. Let's take the continent we chose - **`r params$continent`** - and look if there's a statistical relationship between GDP per capita, life expectancy and population.


```{r}
gapminder_continent <- gapminder %>% 
  filter(continent == params$continent)
cont_lm <- lm(pop ~ lifeExp + gdpPercap, data = gapminder_continent)

 cont_lm %>% 
  summary() %>% 
  xtable::xtable() %>% 
  knitr::kable("latex")
```