---
title: Translating Regex to SQL String Operations
author: Amit Levinson
date: '2022-06-02'
slug: regex-in-sql
categories: [R]
tags: [R, Python, SQL]
subtitle: 'Learning how to do several string operations with SQL'
summary: 'Unfortunately, a lot of Regex operations are not available in (MS)SQL. In this post I show how I take several regex operations I like and convert thme to string manipulations with SQL syntax'
featured: yes
image:
  caption: ''
  focal_point: ''
  preview_only: yes
projects: []
draft: true
codefolding_show: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE)
```
### Setup

Let's start by setting up a local connection both for Python (call it `pyconn`) and R (`rconn`) we'll be using. I'm connecting to a local MSSQL server we'll query moving forward.

##### Python

```{python}
import pandas as pd
import re

payments = pd.read_csv('data.csv')
payments
```

##### R
```{r}
library(tidyverse)

payments <- read_csv('data.csv')
```

```{r echo = FALSE}
library(odbc)
library(DBI)
rconn <- dbConnect(odbc(),
                      Driver = "SQL Server",
                      Server = "localhost\\SQLEXPRESS",
                      Database = "regex")

# Partial strings
partial_identifiers <- data.frame(
  id = c("a2209370-bdb0-44c5-b4c5", "1746-4ea9-9a33-f8cc3dd07b8e", "b510-442c-b0d1", "8742-4396-83ca", "6856-4df8-ac69")
)
# dbWriteTable(rconn, 'Partial_Identifiers', partial_identifiers)


# content/post/regex-in-sql/
payments <- read_csv('data.csv')
```



### Lookarounds 👀 — Extracting the domain from an email address

[lookarounds](https://www.regular-expressions.info/lookaround.html) are definitely one of my favorite and commonly go-to regex operations. As the name implies, a lookaround searches for a pattern and string, while a lookahead searches for the pattern and takes what before it, the lookbehind searches for the pattern and what follows it. They both can be positive or negative, while the former searches for a match (positive match) and the latter searches for the string without a match to the symbol referenced.

Let's try this with a set of emails. **For example you might want to extract the email domains, which is everything that's after the @ symbol.**

##### Python

Returning to our table, we can do this using the following regex operation:

```{python}
(
  payments
  .drop_duplicates()
  .loc[:, ['userId', 'email']]
  .assign(
    emaildomain = lambda df: df.email.str.extract(r'((?<=@).+)') # <- relevant part
    )
  )
```


the pattern `(?<=@).+` essentially extracts any symbol(s) that follow the @ symbol, in this case our domain[^1]. Alternatively if we were interested in extracting email names instead we could use a positive lookahead, looking for the '@' symbol only this time taking what's before it. 

[^1]: The extra parentheses is to solve the 'ValueError: pattern contains no capture groups' error, basically including what is it we want to be captured in our regex.

##### SQL

So how can we do it in SQL? Well, I mainly use it for the positive lookahead/behind, where we can identify the character's location and extract anything after it:

```{sql connection=rconn, echo = TRUE}
SELECT DISTINCT p.userId,
  email,
  RIGHT(EMAIL, LEN(EMAIL) - CHARINDEX('@',EMAIL)) AS email_domain
FROM PAYMENTS p
```


We're leveraging the function `CHARINDEX` in order to identifying the location of the '@' symbol, and then extract all text from that location forward.

### Partial string join — identifying a string from a partial match

I wouldn't say this is a common thing I do, but I had to do it once and was pretty pleased with the solution. Assuming you get a list of only partial unique identifiers, how can we identify the correct observation? 

For example, you received from some partner a list of ids he has for each payment. However, what he has is only a part of the full string, as we can see below:

```{r}
knitr::kable(partial_identifiers)
```
How can we select the relevant payments?

##### R

At the time I encountered this I was using mainly R and solved it with that. Let's solve it first and then go over the solution:

```{r}
library(purrr)
library(stringr)

check_payment_id_exists <- function (payment_id) {
  result = partial_identifiers$id[map_lgl(partial_identifiers$id, ~ str_detect(payment_id, .))]
  result = ifelse(is_empty(result), NA, result)
  return(result)
}

payments %>% 
  mutate(
    identifier = map_chr(payment_identifier, check_payment_id_exists),
    .before = 'payment_identifier', .keep='used'
  )

```

So what do we have here? Well the idea is to iterate across all payment identifiers we have and see which of the partial identifiers matches it. To do this we break the process up, the first section is a function to do exactly that, identify if a payment matches any of the partial identifiers, and the second section is the analysis in which I iterate across all payment identifiers and check if they match. 
<< Explanation >>

##### SQL

```{sql connection=rconn, echo = TRUE}
SELECT payment_identifier,
  pi.id
FROM PAYMENTS p
LEFT JOIN partial_identifiers pi on p.payment_identifier like concat('%', pi.id, '%')
```



The idea is pretty straight forward. We can leverage the `LIKE` operator in a join to do the partial matching for us, matching any payment identifiers to the partial identifiers id.

### Extracting / Separating text & digits.

Occasionally you might encounter values that contain both a string and digits combined, for example payment descriptions, email users, security answers and more. Being able to separate the text from numbers might be a necessary step for cleaning our data and further analysis. Let's see how can we do this on the column payment_description.

```{python echo=TRUE}
(
  payments[['payment_description']]
  .assign(
    name = payments.payment_description.str.extract(r'([^0-9]+)'),
    number = payments.payment_description.str.extract(r'(\d+)'),
)
)
payments[['payment_description']].apply(lambda val: re.split(r'(\d+)',val.payment_description), result_type = 'expand')
```





```{r}
# library(readr)
# dat <- read_csv('content/post/regex-in-sql/data.csv')

# dbWriteTable(conn = rconn,
#             'Payments',
#             dat,
#             overwrite = TRUE)
```


```{r echo = FALSE}
dbDisconnect(rconn)
```

